Catecholamines and Hypoglycemia

### 2024-09-12 Christine Chan/EnVision CF/catecholamines_analysis.qmd
catecholamines analysis updates

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(gtsummary)
library(lmerTest)
library(emmeans)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

- Epinephrine and norepinephrine were log transformed for modeling purposes.

```{r data cleaning}
# Import cleaned data
df <- read.csv("./Christine Chan/EnVision CF/Data_Clean/analysis_dataset.csv",
  na.strings = ""
)
# For this analysis, we only want those with catecholamines
df <- df %>%
  group_by(study_id, Date) %>%
  mutate(catecholamines = any(!is.na(Epinephrine) | !is.na(Norepinephrine))) %>%
  filter(catecholamines) %>%
  ungroup()
# # Time as a factor
df$Timepoint <- factor(df$Timepoint, levels = c(0, 10, 30, 60, 90, 120, 150, 180))
```

@@

```{r}
#| label: tbl-1
#| tbl-cap: Participant characteristics (overall)
df %>%
  group_by(study_id) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(
    redcap_data_access_group, age_visit:ethnicity, cftr_mutation_1,
    cftr_mutation_2
  ) %>%
  tbl_summary(by = redcap_data_access_group, missing_text = "Missing") %>%
  add_overall()
```

@@

```{r}
#| label: tbl-70
#| tbl-cap: BG < 70 at Any Point
#| warning: false
df %>%
  filter(Timepoint %in% c(0, 120, 150, 180)) %>%
  pivot_wider(
    names_from = Timepoint, values_from = c(Epinephrine, Norepinephrine),
    id_cols = c(study_id:cftr_mutation_2, Hypo70)
  ) %>%
  select(Hypo70, age_visit:cftr_mutation_2, Epinephrine_0:Norepinephrine_180) %>%
  tbl_summary(by = Hypo70, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Epinephrine) ~ Timepoint * Hypo70 + (1 | study_id), data = df)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
means %>% gt(caption = "LS Means")
# Plot
emmip(mod, Hypo70 ~ Timepoint, CIs = T, type = "response") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Norepinephrine) ~ Timepoint * Hypo70 + (1 | study_id), data = df)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
means %>% gt(caption = "LS Means")
# Plot
emmip(mod, Hypo70 ~ Timepoint, CIs = T, type = "response") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| label: tbl-60
#| tbl-cap: BG < 60 at Any Point
#| warning: false
df %>%
  filter(Timepoint %in% c(0, 120, 150, 180)) %>%
  pivot_wider(
    names_from = Timepoint, values_from = c(Epinephrine, Norepinephrine),
    id_cols = c(study_id:cftr_mutation_2, Hypo60)
  ) %>%
  select(Hypo60, age_visit:cftr_mutation_2, Epinephrine_0:Norepinephrine_180) %>%
  tbl_summary(by = Hypo60, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Epinephrine) ~ Timepoint * Hypo60 + (1 | study_id), data = df)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
means %>% gt(caption = "LS Means")
# Plot
emmip(mod, Hypo60 ~ Timepoint, CIs = T, type = "response") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Norepinephrine) ~ Timepoint * Hypo60 + (1 | study_id), data = df)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
means %>% gt(caption = "LS Means")
# Plot
emmip(mod, Hypo60 ~ Timepoint, CIs = T, type = "response") +
  theme_bw() +
  theme(legend.title = element_blank())
```

>>>>

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(gtsummary)
library(lmerTest)
library(emmeans)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

- Epinephrine, norepinephrine, and insulin were log transformed for modeling purposes.

```{r data cleaning}
# Import cleaned data
df <- read.csv("./Christine Chan/EnVision CF/Data_Clean/analysis_dataset.csv",
  na.strings = ""
)
# Timepoint and variables lists
timepoints <- c(0, 10, 30, 60, 90, 120, 150, 180)
limited_timepoints <- c(0, 120, 150, 180)
glucoses <- paste0("Glucose_", timepoints)
insulins <- paste0("Insulin_", timepoints)
epinephrines <- paste0("Epinephrine_", timepoints)
norepinephrines <- paste0("Norepinephrine_", timepoints)
# For this analysis, we only want those with catecholamines measured
df$catecholamines <- apply(
  df[, c(epinephrines, norepinephrines)], 1,
  function(r) {
    any(!is.na(r))
  }
)
df <- df %>% filter(catecholamines)
# Make a long dataset for mixed models
lmm_df <- df %>%
  select(
    study_id, Hypo60, Hypo70,
    all_of(c(glucoses, insulins, epinephrines, norepinephrines))
  ) %>%
  pivot_longer(all_of(c(glucoses, insulins, epinephrines, norepinephrines)),
    names_to = c(".value", "Timepoint"),
    names_pattern = "(.*)_(.*)"
  )
```

@@

```{r}
#| label: tbl-1
#| tbl-cap: Participant characteristics (overall)
df %>%
  select(
    redcap_data_access_group, age_visit:ethnicity, cftr_mutation_1,
    cftr_mutation_2
  ) %>%
  tbl_summary(by = redcap_data_access_group, missing_text = "Missing") %>%
  add_overall()
```

@@

```{r}
#| label: tbl-70
#| tbl-cap: BG < 70 at Any Point
#| warning: false
df %>%
  select(
    Hypo70, age_visit:cftr_mutation_2, hypoglycemia_symptoms_yesno,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo70, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

## Glucose

Glucose was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable.

```{r}
#| warning: false
# Fit
mod <- lmer(Glucose ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  ylab("LS Mean Glucose") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

## Insulin

Insulin was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable.

```{r}
#| warning: false
# Fit
mod <- lmer(log(Insulin) ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  ylab("LS Mean Insulin") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Epinephrine) ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Mean Epinephrine") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Norepinephrine) ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Mean Norepinephrine") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| label: tbl-60
#| tbl-cap: BG < 60 at Any Point
#| warning: false
df %>%
  select(
    Hypo60, age_visit:cftr_mutation_2, hypoglycemia_symptoms_yesno,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo60, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

## Glucose

Glucose was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable.

```{r}
#| warning: false
# Fit
mod <- lmer(Glucose ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  ylab("LS Mean Glucose") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

## Insulin

Insulin was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable.

```{r}
#| warning: false
# Fit
mod <- lmer(log(Insulin) ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  ylab("LS Mean Insulin") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Epinephrine) ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Mean Epinephrine") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| warning: false
# Fit
mod <- lmer(log(Norepinephrine) ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Means") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Mean Norepinephrine") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```
